---
title: "Project-Scribble-SS"
author: "Saumya Singh"
date: "7/30/2020"
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(dplyr)
library(faraway)
library(lmtest)
library(MASS)
num_sims = 2000

temp_DF = read_csv("Team-Illinois//data//Consolidated_95_observations.csv") 


calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```

## Variable selection


```{r cars}
data_set = c("Years of Potential Life Lost Rate",
               "% Fair or Poor Health",
               "% Low Birthweight",
               "% Adults with Obesity",
               "% Excessive Drinking",
               "High School Graduation Rate",
               "% Unemployed",
               "Annual Average Violent Crimes",
               "Presence of Water Violation",
               "% Frequent Physical Distress",
               "HIV Prevalence Rate",
               "% Limited Access to Healthy Foods",
               "Median Household Income",
               "% Homeowners")

final_df = subset(temp_DF, select = data_set)
names(final_df) = c("YPLLRate","HealthPoorPct","LBWPct","ObeseAdultsPct","DrinkExcPct",
                    "HSGradRate","UnemployedPct","ViolsCrimeAnlAvg","WaterViol",
                    "DistressFreqPhysPct","HIVRate","FoodHealthLimAccPct",
                    "IncHousehldMedian","HomeownersPct")
final_df = na.omit(final_df)
final_df$`WaterViol` = as.factor(final_df$`WaterViol`)

indexes = sample(nrow(final_df), (nrow(final_df)) /2)
train_df = final_df[indexes,]
test_df = final_df[-indexes,]

```

## Pairs plot of data

```{r, fig.height=10, fig.width=15}
pairs(test_df)
```

## TODO try transformations of predictors

```{r}
par(mfrow = c(2,2))

plot(final_df$HSGradRate, 
     final_df$YPLLRate, 
     col = "dodgerblue", 
     pch = 20,
)

plot(final_df$HSGradRate, 
     log(final_df$YPLLRate), 
     col = "dodgerblue", 
     pch = 20,
)

```


## Model 1: with ALL manually selected predictors 

```{r}
#create a model with all predictors
model_1 = lm(YPLLRate ~ . , data = test_df)

```

**Fitted versus Residuals Plot**
```{r}
plot(fitted(model_1), resid(model_1), col = "grey", pch = 20,
     xlab = "Fitted", ylab = "Residuals", main = " Model 1")
abline(h = 0, col = "darkorange", lwd = 2)


hist(resid(model_1),
     xlab   = "Residuals",
     main   = "Histogram of Residuals, Model 1",
     col    = "darkorange",
     border = "dodgerblue",
     breaks = 45)

```
```{r}
qqnorm(resid(model_1), main = "Normal Q-Q Plot, Model 1", col = "darkgrey")
qqline(resid(model_1), col = "dodgerblue", lwd = 2)
```

**Other diagnostics**

- VIF
```{r}
vif(model_1)
```

- Adjusted R Squared
```{r}
summary(model_1)$adj.r.squared 
```

- LOOCV RMSE
```{r}
calc_loocv_rmse(model_1)
```

- BP Test
```{r}
bptest(model_1)
```


## BOX-COX transformation on response

```{r}
boxcox(model_1, lambda = seq(-0.25, 0.75, by = 0.05), plotit = TRUE)
```

## MODEL 2 - transformed response
```{r}
model_1_cox = lm((((YPLLRate ^ 0.3) - 1) / 0.3) ~ . , data = test_df)
```

**Fitted versus Residuals Plot**
```{r}
plot(fitted(model_1_cox), resid(model_1_cox), col = "grey", pch = 20,
     xlab = "Fitted", ylab = "Residuals", main = " Model 2")
abline(h = 0, col = "darkorange", lwd = 2)
```

**Fitted versus Residuals Plot**
```{r}
hist(resid(model_1_cox),
     xlab   = "Residuals",
     main   = "Histogram of Residuals, Model 2",
     col    = "darkorange",
     border = "dodgerblue",
     breaks = 45)
```

**QQ Plot**
```{r}
qqnorm(resid(model_1_cox), main = "Normal Q-Q Plot, Model 2", col = "darkgrey")
qqline(resid(model_1_cox), col = "dodgerblue", lwd = 2)
```

**Other diagnostics**

- VIF
```{r}
vif(model_1_cox)
```

- Adjusted R Squared
```{r}
summary(model_1_cox)$adj.r.squared 
```

- LOOCV RMSE
```{r}
calc_loocv_rmse(model_1_cox)
```

- BP TEST
```{r}
  bptest(model_1_cox)
```

## BACKWARD AIC on MODEL 1
```{r}
(back_aic_mod = step(model_1, direction = "backward", trace=0))

```

```{r}
vif(back_aic_mod)
```

## BACKWARD AIC on MODEL 1 - remove DistressFreqPhysPct

```{r}
model_3 = lm(formula = (((YPLLRate ^ 0.3) - 1) / 0.3)  ~ HealthPoorPct + LBWPct + ObeseAdultsPct + DrinkExcPct + WaterViol + FoodHealthLimAccPct + IncHousehldMedian + HomeownersPct, data = test_df)
```

- VIF
```{r}
vif(model_3)
```

- Adjusted R Squared
```{r}
summary(model_3)$adj.r.squared 
```

- LOOCV RMSE
```{r}
calc_loocv_rmse(model_3)
```

- BP TEST
```{r}
  bptest(model_3)
```







```{r}
rmse  = function(actual, predicted) {
  sqrt(mean((actual - predicted) ^ 2))
}
```

