---
title: 'Predicting Premature Death Based on Health and Socio-economic Factors'
author: 'Team Illinois: Data Analysis Project Report'
date: '08/07/2020'
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---
***

**Team Illinois Members**

  - Julia Dawson, NETID: juliaed3
  
  - Joseph Stewart, NETID: TODO
  
  - Saumya Singh, NETID: saumyas2
  
***

```{r setup, include=FALSE}

# Read as packages here
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(dplyr)
library(faraway)
library(lmtest)
library(MASS)
library(knitr)
library(kableExtra)
```

# Introduction


# Methods

## Approach 1

## Approach 2

## Approach 3
# Data Cleaning: 
We started with a script that merged two sheets from the original data. The team chose to keep variables that contained rates and to drop race demographics. For this approach we columns with too many NA's (more than our result YPLLRate) and then omitted observations containing 'NA'. We started with 92 variables and 3141 observations and after cleaning were left with 71 variables and 2333 observations. Lastly, several variables were coerced to factors and short names were added.

# Test Statistics used:
**adjusted R-squared** is a modified version of R-squared that has been adjusted for the number of predictors in the model. The adjusted R-squared increases only if the new term improves the model more than would be expected by chance. It decreases when a predictor improves the model by less than expected by chance.

**LOOCV RMSE**: leave-one out cross-validated RMSE allows us to check for overfitting.

**Correlation Coefficients** were used to assess how highly correlated predictors are with the result. They were also used to consider which predictors may be interacting.

**Partial Correlation Coefficients** The partial correlation coefficient was used in predictor selection in addition to correlation because it it controls for confounding variables in a model.

**Shapiro-Wilk Test**: Shapiro test assesses the normality of errors. The null hypothesis assumes data were sampled from a normal distribution. A small p-value rejects the null hypothesis. 

**Breusch-Pagan-Godfrey Test**: <https://www.statisticshowto.com/breusch-pagan-godfrey-test/>
Heteroscedastic data. ... Heteroscedasticity means “differently scattered”; this is opposite to homoscedastic, which means “same scatter.” Homoscedasticity in regression is an important assumption; if the assumption is violated, you won't be able to use regression analysis. If bptest gives a very small p-value so we reject the null hypothesis of constant variance.

**Durbin-Watson**: <https://www.statisticshowto.com/durbin-watson-test-coefficient>
dwtest p — p-value of test p-value of the test, returned as a numeric value. dwtest tests whether the residuals are uncorrelated, against the alternative that autocorrelation exists among the residuals. A small p-value indicates that the residuals are autocorrelated. 0 to < 2 is positive autocorrelation. >2 to 4 is negative autocorrelation. Test statistics between 1.5 and 2.5 are relatively normal. The null hypothesis for dwtest is no first order correlation exists. Our p-value is higher than alpha 0.05 so we cannot reject the null hypothesis.

# Model Search:
*Result selection: We chose YPLLRate as our response variable. YPLLRate had the most highly correlated predictors. YPLL is the years of potential life lost before age 75 per 100,000 population (age-adjusted). YPLLRate is a measure of premature death and is used on an interactive site <https://www.countyhealthrankings.org/explore-health-rankings> to allow people to compare their State and County to others. The statement is on their home page is interesting... _"The annual Rankings provide a revealing snapshot of how health is influenced by where we live, learn, work, and play. They provide a starting point for change in communities."_ . Data is gathered from a variety of sources to create a model for comparison between States and Counties (<https://www.countyhealthrankings.org/explore-health-rankings/measures-data-sources/2020-measures>). Several other variables were considered to be the result but YPLLRate was chosen as the best representative of the data.  

**Predictor selection:** 
  + Predictors were chosen if;
    - they had some correlation with result (YPLLRate).
    - their partial correlation with result (YPLLRate) was high while controlling for the other model predictors.
    - they showed up in several large automatically generated models (full additive, full interactive, backward AIC)
    - they were representative of factor we thought could be significant
    
  + Predictors were eliminated one by one if:
    - they had high p-values in a model
    - removal had good or low affect on adjusted R-squared and LOOCV-RMSE
    - predictor was highly correlated with another predictor in the model
    
  + Automatic Model builds that were used for comparison:
    - Additive model all and backward AIC
    - Multiplicative model all and backward AIC
    - regsubsets all possible models (we tried this but it ran for three days and crashed the computer)
    
**Problems with Model_3 and methods to improve it:**
  + LOOCV RMSE (Cross-Validated RMSE) wa too high (355)
    - Outliers were removed and improved LOOCV RMSE by 30%.
    - Transforming a few predictors with log and higher order terms improved the adjusted-rsquared and LOOCV RMSE.
    - Adding some two-way interactions improved the adjusted r-squared and LOOCV RMSE slightly.
    - BoxCox recommended a lambda of ~0.8. Transformation of response YPLLRate^lambda worked best ($Y^\lambda$ rather than $\frac{(Y^\lambda - 1)}\lambda$). It greatly improved LOOCV RMSE and seemed to improve the normality problems with our response variable. The LOOCV RMSE was well below the standard deviation of YPLLRate. The large tails on the Q-Q plot shrank to almost nothing.
    
  + Model suffers from heteroscedasticity (Breusch-Pagan Test) and correlation between residuals (Durbin Watson Test)     
    - scaled the values of predictors - this did not help
    - weighted each observations - this did not help
    - Remove predictors one by one with highest p-value if have good or little affect. This improved all measures.
    - Removing outliers helped some.
    - Transforming the Response YPLLRate reduced BP value 25% but is still too high.
    - we tried a model that does these automatically like glmnet but the results and model were difficult to interpret and compare in this short timeframe.

```{r, eval=T, echo=F}

#model_1 = lm(Employed ~ . , data = longley)

#Read data for Model 3
Raw95naomit = read.csv(".\\data\\DataCleanedJulia.csv") 
#str(Raw95naomit)
#Force character arrays to be factors.  
Raw95naomit$State = as.factor(Raw95naomit$State)
Raw95naomit$Region= as.factor(Raw95naomit$Region)
Raw95naomit$WaterViol = as.factor(Raw95naomit$WaterViol)
#Make this model to can figure out the lambda
mdl3t = lm(formula = YPLLRate ~ HealthPoorPct + 
             FoodEnvirIX + PhysInactPct + 
             PhysPrimCareRate + 
             MammAnnualPct +   
             Income20thpcntl + IncomeRatio + SocsAssRate + 
             InjuryDeathRate + HousSvrProbPct + 
             HousInadFacil + LifeExpect + 
             DeathAgeAdjRate + VaccinatedPct + 
             IncHousehldMedian + 
             PopGE65Pct + PopNativePct + PopPacificPct + 
             EnglishNotProfPct + PopRuralPct
           + PopGE65Pct:DeathAgeAdjRate
           + HealthPoorPct:PopFemalePct
           + LBWPct:IncHousehldMedian
           , data = Raw95naomit)

#Transform according to BoxCox lambda recommendation
lambda_mdl3 = boxcox(mdl3t, plotit = FALSE) #lambda a little less than 0.8
lmda_mdl3 = lambda_mdl3$x[which.max(lambda_mdl3$y)]

#Detect outliers
cdnaomit = cooks.distance(mdl3t)

#Create Model_3
model_3 = lm(formula = YPLLRate^lmda_mdl3 ~ HealthPoorPct + 
             FoodEnvirIX + PhysInactPct + 
             PhysPrimCareRate + 
             MammAnnualPct +   
             Income20thpcntl + IncomeRatio + SocsAssRate + 
             InjuryDeathRate + HousSvrProbPct + 
             HousInadFacil + LifeExpect + 
             DeathAgeAdjRate + VaccinatedPct + 
             IncHousehldMedian + 
             PopGE65Pct + PopNativePct + PopPacificPct + 
             EnglishNotProfPct + PopRuralPct
           + PopGE65Pct:DeathAgeAdjRate
           + HealthPoorPct:PopFemalePct
           + LBWPct:IncHousehldMedian
           , data = Raw95naomit, 
           subset = cdnaomit <= 4/length(cdnaomit))
```
# The three models are shown below:

**Model_1**
```{r}
formula(model_3)
```

**Model_2**
```{r}
formula(model_3)
```
**Model_3**
```{r}
formula(model_3)
```


# Model Diagnostics

```{r, eval=T, echo=F}

plot_fitted_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange", title) {
  
  plot(fitted(model), resid(model), col = pointcol, pch = 20, cex = 1.5, xlab = "Fitted", ylab = "Residuals", main = title)
  abline(h = 0, col = linecol, lwd = 2)
}

plot_qq = function(model, pointcol = "dodgerblue", linecol = "darkorange", title) {
  qqnorm(resid(model), main = title, col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)
}

plot_hist_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange", title) {
  hist(resid(model),
       xlab   = "Residuals",
       main   = title,
       col    = pointcol,
       border = linecol,
       breaks = 45)
}

```

**Residuals Plot**
```{r fig.height=8, fig.width=20}
par(mfrow = c(1,3))
#plot_fitted_resid(model_1, title="Model 1")
#plot_fitted_resid(model_2, title="Model 2")
plot_fitted_resid(model_3, title="Model 3")

```

**Q-Q Plot**
```{r}
par(mfrow = c(1,3))
plot_qq(model_3, title="Model 3")
```




# Results


# Discussion

```{r, eval=T, echo=F}

get_bp = function(model) {
  bptest(model)$p.value
  #decide = unname(bptest(model)$p.value < alpha)
  #ifelse(decide, "Reject", "Fail to Reject")
}

get_sw = function(model) {
  shapiro.test(resid(model))$p.value
  #decide = unname(shapiro.test(resid(model))$p.value < alpha)
  #ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}

get_rmse = function(model) {
  sqrt(mean(resid(model) ^ 2))
}

get_AIC = function(model) {
  extractAIC(model)[2]
}
#Joe and Saumya - we need to normalize RMSE to compare so each of us needs to supply an stddev and point to your YPLLRate
rmse_norm_1 = get_rmse(model_3)/sd(log(Raw95naomit$YPLLRate)) #Model 1 result transform
info_1 = list(get_bp(model_3), get_sw(model_3), get_num_params(model_3), 
              get_loocv_rmse(model_3), get_adj_r2(model_3), get_rmse(model_3), 
              rmse_norm_1, get_AIC(model_3))

rmse_norm_2 = get_rmse(model_3)/sd((Raw95naomit$YPLLRate^lmda_mdl3 - 1)/lmda_mdl3) #Model 2 result transform
info_2 = list(get_bp(model_3), get_sw(model_3), get_num_params(model_3), 
              get_loocv_rmse(model_3), get_adj_r2(model_3), get_rmse(model_3), 
              rmse_norm_2, get_AIC(model_3))

rmse_norm_3 = get_rmse(model_3)/sd(Raw95naomit$YPLLRate^lmda_mdl3)
info_3 = list(get_bp(model_3), get_sw(model_3), get_num_params(model_3), 
              get_loocv_rmse(model_3), get_adj_r2(model_3), get_rmse(model_3), 
              rmse_norm_3, get_AIC(model_3))

model_info = cbind(info_1, info_2, info_3)

rownames(model_info) = c("BP Test", "Shapiro","No params", "LOOV RMSE", "Adj R2", "RMSE", "norm RMSE", "AIC")
colnames(model_info) = c("Model 1", "Model 2", "Model 3")
knitr::kable(model_info, digits=3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```


# Appendix
